Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-20 manuscript/climate_sensitivity_draft.Rmd
Added feedback up to beginning of last 3 pgrph of results
#65

Height measurements (n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights). Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic, ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Species-specific height allometries were developed (Table S2). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements. **see comments 16,17**

Crown positions were recorded in the field during the growing season of 2018 following the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed. As it was unfeasible to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years. While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig. 2d) and height variance since the beginning of the study period indicated that change was likely slow (Fig. S3). **see comment 18,19** Specifically, average tree height growth was confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.

@@

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with widespread reductions in tree growth and anomalously dry peak growing season climatic conditions. Specifically, candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($R$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R. **see comment 24**
Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the current-year months to which annual growth was most sensitive for trees at this site. We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration (PET) and precipitation (PRE). These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014). The driest years were identified through ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest. **see comment 25**

@@

The droughts differed in intensity and prior onset (Table S3, Fig. S2). In terms of intensity during the peak growing season of the focal year, the 1999 drought was the most intense (lowest PDSI) during May-July. It was surpassed in intensity in August by the 1966 drought, which was otherwise the second most intense during the peak growing season. The 1977 was the least intense throughout the growing season. All droughts increased in intensity from May through August, but differed in the months previous. The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall. Similarly, the 1999 drought was preceded by severe to extreme drought starting the previous fall, but before that conditions were wetter than average until the previous June. The 1977 drought was preceded by 2.5 years of near-normal conditions, and was therefore the mildest of the three droughts by any measure.

Community-level tree growth responses to these droughts were modest, with mean resistance values of 0.86, 0.84, and 0.86 for 1966, 1977, and 1999, respectively (Fig. 1b). In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): 29.43% in 1966, 31.86% in 1977, and 26.81% in 1999. Some trees exhibited increased growth: ($R$ > 1.0): 26.31% in 1966, 21.92% in 1977, and 25.57% in 1999.

@@

Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015] (*H1.0*), although there was no signficant effect for 1977 or 1999 individually (Tables 1, 4). The same held true for $ln[H]$ as a univariate predictor (*H1.1*; Tables 1, 4). When combined with other predictor variariables in our multivariate models, the top models usually included an effect of $ln[H]$, and its coefficient was consistently negative, as predicted (Tables 1, 5). We note that a non-signficant positive correlation between $ln[H]$ and $R$ for 1999 became negative in the context of the multivariate models, again supporting *H1.1* (Table 1).

Crown position was generally correlated with $H$, but with substantial variation (Fig. 2d). Crown position was a much poorer predictor of $R$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1). When considered alone, $CP$ had a significant influence only in the 1966 drought, during which trees with dominant $CP$ had the lowest $R$. When considered in conjunction with $H$, $CP$ came out as a significant predictor only for the 1999 drought, during which suppressed and then intermediate trees had the lowest $R$. Crown position was included in almost half of the top models, with mixed results as to how $R$ varied with $CP$ (Table 5). Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees. Thus, $CP$ was sometimes a useful predictor of $R$, but overall had a weak effect relative to that of $H$.

In the non-drought years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig. 2). Specifically, maximum daily wind speeds were signficantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig. 2a). Relative humidity was also somewhat lower during June-August, ranging from ~50-80 above the canopy and ~60-90% in the understory (Fig. 2b). Air temperature did not vary across the vertical profile (Fig. 2c).

Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the idea that trees in moist microsites would suffer less during drought. Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction (*H1.3*), which could indicate that smaller trees (with smaller rooting volume) have a greater tendency to suffer more in drier microenvironments with greater depth to the water table. *H1.3* was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).

@@

We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5). Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Table 4), and therefore these were excluded as candidate variables for the full multivariate models. In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4). Of these, $PLA_{dry}$ was a strong predictor for the 1966 drought and overall, with consistently negative coefficients across all droughts. $\pi_{tlp}$ never came out as significant (dAIC $\le$ 2) in the univariate models but had a consistently negative coefficient (Table 4). Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977. Results were similar in the context of multivariate models (Table 5), except that $PLA_{dry}$ had a small positive coefficient in the one 1999 model in which it was included. $PLA_{dry}$, and $\pi_{tlp}$ were each included in the top overall model and in the top model for two of the three individual droughts (Table 5).

We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance. In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4). However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the cases of $WD$ and $LMA$, which were not useful predictors of $R$. Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig. 2e-f).

We support the hypothesis (*H2.3*) that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Tables 1,5). As discussed above, there was little meaningful variation in traits with height at the community level. When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $R$ was consistently negative (Table 5)--reversing a non-signficant positive $ln[H]$-$R$ correlation in the univariate model for the 1999 drought (Table 4).

>>>>

Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights). Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic, ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H]$). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements. **see comments 16,17**

Crown positions ($CP$) were recorded in the field during the growing season of 2018 following the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed. As it was unfeasible to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years. While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig. 2d) and height variance since the beginning of the study period indicated that change was likely slow (Fig. S3). **see comment 18,19** Specifically, average tree height growth was confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.

@@

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with widespread reductions in tree growth and anomalously dry peak growing season climatic conditions. Specifically, candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($R$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R. Specifically, $R$ values $\le$ 0.7 indicated growth reductions, whereas values $\ge$ 1.0 indicated increased growth. **see comment 24**
Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the current-year months to which annual growth was most sensitive for trees at this site. We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration (PET) and precipitation (PRE). These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014). The driest years were identified through ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest. **see comment 25**

@@

The droughts differed in intensity and prior onset (Table S3, Fig. S2). 1999 for example was the most intense (having the lowest PDSI) during May-July, while 1966 had a more intense August PDSI and overall had the second strongest drought during the growing season. The 1977 drought was the least intense throughout the growing season. All droughts increased in intensity from May through August, but differed in the previous months. The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall. Similarly, the 1999 drought was preceded by severe to extreme drought starting the previous fall, but before that conditions were wetter than average until the previous June. The 1977 drought was preceded by 2.5 years of near-normal conditions, and was therefore the mildest of the three droughts by any measure.

Community-level tree growth responses to these droughts were modest, with mean resistance values of 0.86, 0.84, and 0.86 for 1966, 1977, and 1999, respectively (Fig. 1b). All droughts yielded about 5% more trees experiencing negative growth ($R$ $\le$ 0.7) as defined by our drought threshold, specifically, 29.43% in 1966, 31.86% in 1977, and 26.81% in 1999. However, some trees exhibited increased growth: ($R$ > 1.0): 26.31% in 1966, 21.92% in 1977, and 25.57% in 1999. **see comment 33**

@@

Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015] (*H1.0*), although there was no significant effect during 1977 or 1999 individually (Tables 1, 4). The same held true for $ln[H]$ as a univariate predictor (*H1.1*; Tables 1, 4). When combined with other predictor variariables in our multivariate models, the top models usually included an effect of $ln[H]$, and its coefficient was consistently negative, as predicted (Tables 1, 5). We note that a non-signficant positive correlation between $ln[H]$ and $R$ for 1999 became negative in the context of the multivariate models, again supporting *H1.1* (Table 1).

Crown position was generally correlated with $H$, but with substantial variation (Fig. 2d). Crown position was a much poorer predictor of $R$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1). **see comment 34** When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $R$. When considered in conjunction with $H$, $CP$ resulted as a significant predictor only for the 1999 drought, during which suppressed and then intermediate trees had the lowest $R$. Crown position was included in almost half of the top models, with mixed results as to how $R$ varied with $CP$ (Table 5). Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees. Although $CP$ was a useful predictor of $R$, it overall had a weak effect relative to that of $H$.

In the non-drought years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig. 2). Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig. 2a). Relative humidity was also somewhat lower during June-August, ranging from ~50-80 above the canopy and ~60-90% in the understory (Fig. 2b). Air temperature did not vary across the vertical profile (Fig. 2c).

Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the hypothesis that trees in moist microsites would suffer less during drought. Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction (*H1.3*), which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments that have a deeper water table. *H1.3* was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).

@@

Our results partially support Hypothesis *H2.1*, where species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Tables 1,4,5). Therefore, these were excluded as candidate variables for the full multivariate models. In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4). $PLA_{dry}$ was a strong predictor for the 1966 drought and all droughts combined, with consistently negative coefficients. $\pi_{tlp}$ was not significant in the univariate models but had a consistently negative coefficient (Table 4). Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977. Results were similar in the context of multivariate models (Table 5), except that $PLA_{dry}$ had a small positive coefficient in the one 1999 model in which it was included. $PLA_{dry}$, and $\pi_{tlp}$ were each included in the top overall model and in the top model for two of the three individual droughts (Table 5).

The results reject Hypothesis *H2.2*, finding no evidence that taller trees had traits associated with greater drought resistance. In part because of the large sample size (n=7549 trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4).**see comment 35** $WD$ and $LMA$ were the only traits where the correlation matched the predicted direction (*i.e.*, evidence that more drought-resistant traits are associated to taller trees), but ultimately neither of these traits were useful predictors of $R$. Furthermore, although correlations were statistically significant, trait variability within each height class overwhelmed any vertical trends (Fig. 2e-f).

Our findings support the hypothesis *H2.3* that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Tables 1,5). As discussed above, there was little meaningful variation in traits with height at the community level. When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $R$ was consistently negative (Table 5). Simultaneously in the 1999 univariate model, this co-consideration resulted in a non-signficant positive $ln[H]$-$R$ correlation (Table 4).
